---
title: "Review of Connecticut's Lake systems focusing in geography, aquatic life, and regional impact."
author: "Maddy Shantz, Abby Nolan, Sophia Gee"
format: typst
editor: visual
bibliography: references.bib
echo: false
---

```{r}
options(styler.save_after_styling = TRUE)
library(tidyverse)
library(naniar)
```

## Introduction

### Background and Data Origin

The individuals under study in this data set are land-enclosed bodies of water in CT. distinct(six_lakes, Name) distinct(twelve_lakes, Name). There are 152 individual land-enclosed bodies of water that we will be using for this project. count(six_lakes, Name) count(twelve_lakes, Name). There are repeated measurements on the land-enclosed bodies of water in CT between the 2006 and 2012 datasets, intersect(names(six_lakes), names(twelve_lakes).

The dataset contains both quantitative and categorical measures. The quantitative variables are the acres, shape_area, and shape_length. The categorical variables are the object ID, 305B ID, Name, Location, Aquatic Life, Recreation, fish consumption, cycle, and drinking water supply. The variables that we are specifically looking at are the Name, Cycle, Aquatic Life, and Drinking Water Supply, which are all categorical variables.

The variables that are of interest to us are the aquatic life, drinking water supply, location, shape_area and shape_length. The aquatic life and drinking water supply variables are essential to the first question that asks about the correlation of the two variables during the year 2006 and 2012. We want to see if there is a correlation between the water being potable, and there being a dense aquatic life presence. We then want to deepen our initial question by seeing if the location variable has any impact on the drinking water supply and the aquatic life variables. For example, if the water is located in Bridgeport versus Ridgefield, does that impact the outcome of the first question.

### Literature Review

#### Concerns and Threats of Contamination on Aquatic Ecosystems

@bashir2020

**Brief Summary:** The book "Bio-remediation and Biotechnology: Sustainable Approaches to Pollution Degradation" includes a chapter entitled 'Concerns and Threats of Contamination on Aquatic Ecosystems' that defined water contamination as the outcome associated with human activities like urbanization, industrialization, and agricultural activities. Specific contaminants are from the overuse of pesticides, fertilizers, and sewage from industrial and residential areas. There is a need for regular monitoring and controlling of pollutant discharge into the nearby aquatic environs.

#### State of the World's Freshwater Ecosystems: Physical, Chemical, and Biological Changes

@carpenter2011

In the "Annual Review of Environment and Resources" the section 'State of the World's Freshwater Ecosystems: Physical, Chemical, and Biological Changes' discusses the large-scale changes in freshwater. This includes physical, chemical and biological are associated with human factors, yet there is a lack of addressing the institutional aspects of water management. This paper aims to discuss the natural science aspects of freshwater resources while trying to emphasize the comprehensive approach that is needed to sustaining freshwater resources.

### Research Goals and Questions

Our main research goals were to explore the correlation of aquatic life support and drinking water supply in the years 2006 and 2012, determine if there is a regional impact on the aquatic life support and drinking water supply correlation, and determine how have the length and area of the lakes changed over time.

This data is located on the Connecticut Open Data website and the Department of Energy and Environmental Protection originally collected this data to assess rivers, lakes, and estuaries in 2006 and 2012 to comply with sections 305(b) and 303(d) of the federal Clean Water Act. Section 305(b) of the Clean Water Act "requires each state to monitor, assess and report on the quality of its waters relative to attainment of designated uses established by the State's water quality standards". Section 303(d) "requires each State to compile a subset of that list identifying only those waters not meeting water quality standards and assign a Total Maximum Daily Load (TMDL) priority ranking to each impaired waterbody". The scale of the data is compiled at a 1:24,000 scale.

## Initial Hypotheses:

Provide a list of detailed hypotheses. For example, you might believe that a certain variable is associated with poverty — or that this association is only mediated through a third variable. Be as precise as possible, stating your hypotheses in terms of population quantities such as means, proportions, correlations, or model parameters.

1.  How are aquatic life support and drinking water supply correlated in the years 2006 and 2012?

- H₀: ρ=0\rho = 0ρ=0
- There is no association between aquatic life support and drinking water supply among Connecticut lakes.
- Hₐ: ρ≠0\rho \neq 0ρ=0
- There is an association between aquatic life support and drinking water supply among Connecticut lakes.

2.  Do the regions where the lakes are located have an impact on the aquatic life support and drinking water supply correlation?

- H₀: ρ=0\rho = 0ρ=0 There is no association between the regions where the lakes are located and the impact on the aquatic life support and drinking water supply correlation.

- Hₐ: ρ≠0\rho \neq 0ρ=0

- There is an association between the regions where the lakes are located and the impact on the aquatic life support and drinking water supply correlation.

3.  How have the length and area changed over time?

## Exploratory Data Analysis:

The data needed to be tidied and clean as we chose a 'messy' data set.

```{r}
twelve_lakes <- read_csv("2012.csv")
six_lakes    <- read_csv("2006.csv")

common_cols <- intersect(names(six_lakes), names(twelve_lakes))

six_lakes_aligned   <- six_lakes |> select(all_of(common_cols))
twelve_lakes_aligned <- twelve_lakes |> select(all_of(common_cols))

total_lakes <- bind_rows(six_lakes_aligned, twelve_lakes_aligned) |> 
  arrange(Cycle)
total_lakes
```

### Question 1: How are aquatic life support and drinking water supply correlated in the years 2006 and 2012?

```{r}
table(total_lakes$`Aquatic Life`)
```

```{r}
table(total_lakes$`Drinking Water Supply`)
```

```{r}
lakes_cleaned <- total_lakes |>
  filter(Cycle %in% c(2006, 2012)) |>
  mutate(
    aquatic_numeric = case_when(
      `Aquatic Life` == "Fully Supporting" ~ 5,
      `Aquatic Life` == "Not Supporting" ~ 4,
      `Aquatic Life` == "Not Assessed" ~ 3,
      `Aquatic Life` == "Not Evaluated In 2006" ~ 2,
      `Aquatic Life` == "Water Not Classified For Drinking Supply" ~ 1,
      TRUE ~ NA_real_
    ),
    drinking_numeric = case_when(
      `Drinking Water Supply` == "Fully Supporting" ~ 5,
      `Drinking Water Supply` == "Not Supporting" ~ 4,
      `Drinking Water Supply` == "Not Assessed" ~ 3,
      `Drinking Water Supply` == "Not Evaluated In 2006" ~ 2,
      `Drinking Water Supply` == "Water Not Classified For Drinking Supply" ~ 1,
      TRUE ~ NA_real_
    )
    ) |>
  filter(!is.na(aquatic_numeric) & !is.na(drinking_numeric))

lakes_wide <- lakes_cleaned |>
  pivot_wider(
    id_cols = `305B ID`,
    names_from = Cycle,
    values_from = c(aquatic_numeric, drinking_numeric)
    ) |>
      drop_na()

cor_matrix <- lakes_wide |>
  select(
    aquatic_numeric_2006, 
    drinking_numeric_2006, 
    aquatic_numeric_2012, 
    drinking_numeric_2012
  ) |>
  cor(method = "spearman")

n_lakes <- nrow(lakes_wide)

print(paste("Number of matched lakes (n):", n_lakes))
print(cor_matrix)

lakes_cleaned |>
  ggplot(aes(x = drinking_numeric, y = aquatic_numeric, color = as.factor(Cycle))) +
  geom_jitter(alpha = 0.5, width = 0.18, height = 0.18) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("2006" = "#56B4E9", "2012" = "#E69F00")) +
  labs(
    title = "Comparison of Trends: 2006 vs 2012",
    x = "Drinking Water Supply Score",
    y = "Aquatic Life Support Score",
    color = "Evaluation Year"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

We fail to reject the null hypothesis claiming there is a correlation between aquatic life and drinking water supply. There is no statistical significant change in the correlation between lake aquatic life support and drinking water supply during the years 2006 and 2012 in Connecticut lakes as the p-value is greater than 0.05.

### Question 2: Do the congressional districts where the lakes are located have an impact on the aquatic life support and drinking water supply correlation?

```{r}
total_lakes_regions <- lakes_cleaned |>
  separate_wider_regex(
    cols = Name,
    patterns = c(name = ".*?", " \\(", place = ".*?", "\\)"),
    too_few = "align_start"
  )
total_lakes_regions

water_impact <- total_lakes_regions |>
  group_by(place, Cycle) |>
  summarise(
    mean_aquatic = mean(aquatic_numeric, na.rm = TRUE),
    mean_drinking = mean(drinking_numeric, na.rm = TRUE),
    .groups = "drop"
  )
water_impact
```

```{r}
district_1 <- water_impact |> 
  filter(place %in% c(
    "Bristol", "Colebrook", "Hartland", "Manchester", "Newington", "Portland", "Winchester"),
  Cycle %in% c(2006, 2012))

district_2 <- water_impact |> 
  filter(place %in% c("Ashford", "Bozrah", "Brooklyn", "Chester", "Colchester", "Columbia", "Coventry", "Durham", "East Haddam", "East Hampton", "East Lyme", "Griswold", "Haddam", "Hebron", "Killingly", "Lebanon", "Ledyard", "Mansfield", "Marlborough", "North Stonington", "Norwich", "Plainfield", "Preston", "Somers", "Sprague", "Sterling", "Thompson", "Tolland", "Union", "Vernon", "Voluntown", "Woodstock"
  ),
  Cycle %in% c(2006, 2012))

district_3 <- water_impact |>
  filter(place %in% c("Branford", "Derby", "Guilford", "Hamden", "New Haven", "North Branford", "Seymour", "Stratford", "Wallingford"),
  Cycle %in% c(2006, 2012)
  )

district_4 <- water_impact |>
  filter(place %in% c("Bridgeport", "Fairfield", "Greenwich", "Ridgefield", "Wilton"),
  Cycle %in% c(2006, 2012))

district_5 <- water_impact |>
  filter(place %in% c("Canaan", "Cheshire", "Cornwall", "Danbury", "Goshen", "Kent", "Meriden", "New Fairfield", "Newtown", "Norfolk", "Plymouth", "Salisbury", "Sharon", "Simsbury", "Torrington", "Watertown", "Wolcott", "Woodbury"),
  Cycle %in% c(2006, 2012))

cong_dist_master <- bind_rows(
  "district_1" = district_1,
  "district_2" = district_2,
  "district_3" = district_3,
  "district_4" = district_4,
  "district_5" = district_5,
  .id = "group"
)
cong_dist_master

c_d_plot <- cong_dist_master |>
  group_by(group, Cycle, place) |> 
  summarise(
    avg_aquatic = mean(mean_aquatic, na.rm = TRUE),
    avg_drinking = mean(mean_drinking, na.rm = TRUE), .groups = "drop")

cong_dist_plot1 <- ggplot(c_d_plot, aes(x = avg_drinking, y = avg_aquatic, color = group)) +
  geom_point(size = 3, alpha = 0.7, na.rm = TRUE) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2, na.rm = TRUE) +
  facet_wrap(~ Cycle) +
  scale_color_discrete(
    labels = c(  "district_1" = "District 1",
                 "district_2" = "District 2",
                 "district_3" = "District 3",
                 "district_4" = "District 4",
                 "district_5" = "District 5")
  ) +
  labs(
    title = "Aquatic Life vs. Drinking Water Correlation by District",
    subtitle = "Varying line slopes indicate districts impact the correlation differently",
    x = "Mean Drinking Water Supply Score",
    y = "Mean Aquatic Life Support Score",
    color = "Congressional District"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12, face = "bold"), 
    legend.position = "bottom"
  )
cong_dist_plot1
```

In Congressional Districts 1, 3, and 5, the steepness of the lines suggests a song and positive relationship, as being almost identical in 2006 and 2012. District 4 has a lower baseline for aquatic life support, but it's slope is similar to Districts 1, 3 and 5.

The major visual difference is in District 2, where in 2006 the line is much steeper, yet gets cut off short. This is likely due to there being no lakes that scored above around a 2.5 in the district on the Drinking Water Supply axis. In 2012, the line flattens out, appearing much less steep than the other districts within 2012.

This graph visually suggests that the region does not heavily impact the correlation, except for the small restriction of range in District 2. The positive relationship between Aquatic Life and Drinking Water Supply behaves the same way regardless of which congressional district the lake is.

### Question 3: How have the length and area changed over time?

```{r}
lakes_change <- six_lakes_aligned |>
   left_join(
      twelve_lakes_aligned,
      by = "305B ID",
      suffix = c("_2006", "_2012")
   )
lakes_change <- lakes_change |> 
select(Name_2006, Name_2012, Shape__Length_2006, Shape__Length_2012, Shape__Area_2006, Shape__Area_2012) |>
mutate(length_ratio = Shape__Length_2012 / Shape__Length_2006,area_ratio = Shape__Area_2012 / Shape__Area_2006)
```

```{r}
lakes_change <- lakes_change |>
  mutate( length_percent_change = (length_ratio - 1) * 100, area_percent_change = (area_ratio - 1) * 100)
```

```{r}
ggplot(lakes_change, aes(x = length_percent_change)) +
  geom_histogram(bins = 30, fill = "steelblue") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs (x = "Percent change in length", y = "Number of lakes", title = "Distribution of Changes in Lake Length from 2006 to 2012") +
  theme_minimal()
```

```{r}
ggplot(lakes_change, aes(x = area_percent_change)) +
  geom_histogram(bins = 30, fill = "steelblue") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs( x = "Percent change in area", y = "Number of lakes", title = "Distribution of Changes in Lake Area from 2006 to 2012") +
  theme_minimal()
```

```{r}
t.test(
  lakes_change$Shape__Length_2012,
  lakes_change$Shape__Length_2006,
  paired = TRUE)
```

```{r}
t.test(
  lakes_change$Shape__Area_2012,
  lakes_change$Shape__Area_2006,
  paired = TRUE)
```

We fail to reject the null hypothesis that there is no change in the length and area of the lakes between 2006 and 2012. There is a significant amount of change in lake length between 2006 and 2012. However, the change in area is not statistically significant. 

## Statistical Methods, Modeling, and Results:

Include your formal statistical analyses such as a formal hypothesis test and modeling. Include your assessment of the assumptions of your methods, with necessary adjustments.

## Data-driven Questions:

What new questions (or hypotheses) have you developed as you explored the data? Provide some description of things that you found interesting when you were looking at the data.

## Discussion:

A brief discussion of your results and how they may be placed in the context of the literature.

## References:

\*She prefers APA

Bashir, I., Lone, F. A., Bhat, R. A., Mir, S. A., Dar, Z. A., & Dar, S. A. (2020). Concerns and Threats of Contamination on Aquatic Ecosystems. Bioremediation and Biotechnology: Sustainable Approaches to Pollution Degradation, 1–26. https://doi.org/10.1007/978-3-030-35691-0_1

Stephen R. Carpenter, Emily H. Stanley, M. Jake Vander Zanden. 2011. State of the World's Freshwater Ecosystems: Physical, Chemical, and Biological Changes. Annual Review of Environment and Resources 36:75-99. https://doi.org/10.1146/annurev-environ-021810-094524

[@carpenter2011] [@bashir2020a]

## Appendix:

Include whatever you feel needs to be included but doesn’t fit in with the above. This could include additional R code not included in the main text. Include a detailed enough description of how you got your data set so someone else could find the same date set. Give web links and/or instructions.
