---
title: "Group Presentation"
author: "Maddy Shantz, Abby Nolan, and Sophia Gee"
format: pptx
editor: visual
echo: false
---

```{r}
twelve_lakes <- read_csv("2012.csv")
six_lakes    <- read_csv("2006.csv")

common_cols <- intersect(names(six_lakes), names(twelve_lakes))

six_lakes_aligned   <- six_lakes |> select(all_of(common_cols))
twelve_lakes_aligned <- twelve_lakes |> select(all_of(common_cols))

total_lakes <- bind_rows(six_lakes_aligned, twelve_lakes_aligned) |> 
  arrange(Cycle)
total_lakes
```

## Introduction

### Data Information

1.  Individuals are land-enclosed bodies of water in CT.

```{r}
distinct(six_lakes_aligned, Name)
distinct(twelve_lakes_aligned, Name)
```

1.  152 individual land-enclosed bodies of water that we will be using for this project.

```{r}
count(six_lakes_aligned, Name) 
count(twelve_lakes_aligned, Name)
```

1.  Yes, there are repeated measurements on the land-enclosed bodies of water in CT between the 2006 and 2012 datasets.

```{r}
intersect(names(six_lakes_aligned), names(twelve_lakes_aligned))
```

1.  The variables that are of interest to us are the aquatic life, drinking water supply, location, shape_area and shape_length. The aquatic life and drinking water supply variables are essential to the first question that asks about the correlation of the two variables during the year 2006 and 2012. We want to see if there is a correlation between the water being potable, and there being a dense aquatic life presence. We then want to deepen our initial question by seeing if the location variable has any impact on the drinking water supply and the aquatic life variables. For example, if the water is located in Bridgeport versus Ridgefield, does that impact the outcome of the first question.

#### Concerns and Threats of Contamination on Aquatic Ecosystems

@bashir2020

- Water contamination is the outcome associated with human activities like urbanization, industrialization, and agricultural activities.

- Specific contaminants are from the overuse of pesticides, fertilizers, and sewage from industrial and residential areas.

- There is a need for regular monitoring and controlling of pollutant discharge into the nearby aquatic environs.

#### State of the World's Freshwater Ecosystems: Physical, Chemical, and Biological Changes

@carpenter2011

- The large-scale changes in freshwater includes physical, chemical and biological aspects that are associated with human factors, yet there is a lack of addressing the institutional aspects of water management.

- This paper aims to discuss the natural science aspects of freshwater resources while trying to emphasize the comprehensive approach that is needed to sustaining freshwater resources.

## Initial Hypotheses

Make them specific and connect to measures in your data set. Where appropriate, write these as formal tests of hypotheses.

1.  How are aquatic life support and drinking water supply correlated in the years 2006 and 2012?

H₀: ρ=0\rho = 0ρ=0

There is no association between aquatic life support and drinking water supply among Connecticut lakes.

Hₐ: ρ≠0\rho \neq 0ρ=0

There is an association between aquatic life support and drinking water supply among Connecticut lakes.

2.  Do the regions where the lakes are located have an impact on the aquatic life support and drinking water supply correlation?

H₀: ρ=0\rho = 0ρ=0 There is no association between the regions where the lakes are located and the impact on the aquatic life support and drinking water supply correlation.

Hₐ: ρ≠0\rho \neq 0ρ=0

There is an association between the regions where the lakes are located and the impact on the aquatic life support and drinking water supply correlation.

3.  How have the length and area changed over time?

## Exploratory Data Analysis

```{r}
options(styler.save_after_styling = TRUE)
library(tidyverse)
install.packages("naniar")
library(naniar)

twelve_lakes <- read_csv("2012.csv")
six_lakes    <- read_csv("2006.csv")

common_cols <- intersect(names(six_lakes), names(twelve_lakes))

six_lakes_aligned   <- six_lakes |> select(all_of(common_cols))
twelve_lakes_aligned <- twelve_lakes |> select(all_of(common_cols))

total_lakes <- bind_rows(six_lakes_aligned, twelve_lakes_aligned) |> 
  arrange(Cycle)
total_lakes
```

## Statistical Methods, Modeling, and Results

### Question 1: How are aquatic life support and drinking water supply correlated in the years 2006 and 2012?

```{r}
table(total_lakes$`Aquatic Life`)
```

```{r}
table(total_lakes$`Drinking Water Supply`)
```

```{r}
lakes_mutated <- total_lakes |>
  filter(Cycle == 2006 | Cycle == 2012) |>
  mutate(
    across(c(`Aquatic Life`, `Drinking Water Supply`), ~ case_when(
      .x == "Fully Supporting" ~ 5,
      .x == "Not Supporting" ~ 4,
      .x == "Not Assessed" ~ 3,
      .x == "Not Evaluated In 2006" ~ 2,
      .x == "Water Not Classified For Drinking Supply" ~ 1
    ))
  )
lakes_mutated

lake_correlations <- lakes_mutated |>
  group_by(`Cycle`) |>
  summarize(correlation = if(sd(`Aquatic Life`, na.rm = TRUE) > 0 & sd(`Drinking Water Supply`, na.rm = TRUE) > 0) {
    cor(`Aquatic Life`, `Drinking Water Supply`, use = "complete.obs")
  } else {
    NA_real_
  }
  )
print(lake_correlations)
```

### Question 2: Do the towns where the lakes are located have an impact on the aquatic life support and drinking water supply correlation?

```{r}
total_lakes_regions <- lakes_mutated |>
  separate_wider_regex(
    cols = Name,
    patterns = c(name = ".*?", " \\(", place = ".*?", "\\)"),
    too_few = "align_start"
  )
total_lakes_regions
```

```{r}
water_impact <- total_lakes_regions |>
  group_by(place) |>
  summarise(
    mean_aquatic = mean(`Aquatic Life`, na.rm = TRUE),
    mean_drinking = mean(`Drinking Water Supply`, na.rm = TRUE)
  )
water_impact
```

```{r}
district_1 <- water_impact |> 
  filter(place %in% c(
    "Bristol", "Colebrook", "Hartland", "Manchester", "Newington", "Portland", "Winchester"),
    Cycle %in% c(2006, 2012))

district_2 <- water_impact |> 
  filter(place %in% c("Ashford", "Bozrah", "Brooklyn", "Chester", "Colchester", "Columbia", "Coventry", "Durham", "East Haddam", "East Hampton", "East Lyme", "Griswold", "Haddam", "Hebron", "Killingly", "Lebanon", "Ledyard", "Mansfield", "Marlborough", "North Stonington", "Norwich", "Plainfield", "Preston", "Somers", "Sprague", "Sterling", "Thompson", "Tolland", "Union", "Vernon", "Voluntown", "Woodstock"
  ),
  Cycle %in% c(2006, 2012))

district_3 <- water_impact |>
  filter(place %in% c("Branford", "Derby", "Guilford", "Hamden", "New Haven", "North Branford", "Seymour", "Stratford", "Wallingford"),
         Cycle %in% c(2006, 2012)
  )

district_4 <- water_impact |>
  filter(place %in% c("Bridgeport", "Fairfield", "Greenwich", "Ridgefield", "Wilton"),
         Cycle %in% c(2006, 2012))

district_5 <- water_impact |>
  filter(place %in% c("Canaan", "Cheshire", "Cornwall", "Danbury", "Goshen", "Kent", "Meriden", "New Fairfield", "Newtown", "Norfolk", "Plymouth", "Salisbury", "Sharon", "Simsbury", "Torrington", "Watertown", "Wolcott", "Woodbury"),
         Cycle %in% c(2006, 2012))

cong_dist_master <- bind_rows(
  "district_1" = district_1,
  "district_2" = district_2,
  "district_3" = district_3,
  "district_4" = district_4,
  "district_5" = district_5,
  .id = "group"
)
cong_dist_master

c_d_plot <- cong_dist_master |>
  group_by(group, Cycle, place) |> 
  summarise(
    avg_aquatic = mean(mean_aquatic, na.rm = TRUE),
    avg_drinking = mean(mean_drinking, na.rm = TRUE), .groups = "drop")

c_d_plot

cong_dist_plot1 <- ggplot(c_d_plot, aes(x = avg_drinking, y = avg_aquatic, color = group)) +
  geom_point(size = 3, alpha = 0.7, na.rm = TRUE) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2, na.rm = TRUE) +
  facet_wrap(~ Cycle) +
  labs(
    title = "Aquatic Life vs. Drinking Water Correlation by District",
    subtitle = "Varying line slopes indicate districts impact the correlation differently",
    x = "Mean Drinking Water Supply Score",
    y = "Mean Aquatic Life Support Score",
    color = "Congressional District"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12, face = "bold"), # Enhances panel titles (2006 vs 2012)
    legend.position = "bottom"
  )
cong_dist_plot1
```

### Question 3: What lake has changed the most regarding length and area in the years of 2006 to 2012?

### How have the length and area changed over time?

```{r}
lakes_change <- six_lakes_aligned |>
  left_join(
    twelve_lakes_aligned,
    by = "305B ID",
    suffix = c("_2006", "_2012")
  )
lakes_change <- lakes_change |> 
  select(Name_2006, Name_2012, Shape__Length_2006, Shape__Length_2012, Shape__Area_2006, Shape__Area_2012) |>
  mutate(length_ratio = Shape__Length_2012 / Shape__Length_2006,area_ratio = Shape__Area_2012 / Shape__Area_2006)
```

## Data-driven Questions

#### add here

## Discussion

#### add here

## References

Bashir, I., Lone, F. A., Bhat, R. A., Mir, S. A., Dar, Z. A., & Dar, S. A. (2020). Concerns and Threats of Contamination on Aquatic Ecosystems. Bioremediation and Biotechnology: Sustainable Approaches to Pollution Degradation, 1–26. https://doi.org/10.1007/978-3-030-35691-0_1

Stephen R. Carpenter, Emily H. Stanley, M. Jake Vander Zanden. 2011. State of the World's Freshwater Ecosystems: Physical, Chemical, and Biological Changes. Annual Review of Environment and Resources 36:75-99. https://doi.org/10.1146/annurev-environ-021810-094524

## Appendix

#### maybe
