---
title: "Review of Connecticut's Lake systems focusing on geography, aquatic life, and regional impact"
author: Madeline Shantz, Abigail Nolan, Sofia Gee
format: 
  revealjs:
    theme: moon
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r}
library(tidyverse)
library(ggplot2)
library(ggthemes)
library(gt)
```

## Introduction to the data

- This data is from the Connecticut Open Data website

- Individuals under study are land-enclosed bodies of water in CT

- Repeated measurements on 152 land-enclosed bodies of water between the 2006 and 2012 data sets

- The dataset contains both quantitative and categorical measures

- The variables of interest are the aquatic life support, drinking water supply, location, shape_area and shape_length

## Literature Review

- 'Concerns and Threats of Contamination on Aquatic Ecosystems' from "Bio-remediation and Biotechnology: Sustainable Approaches to Pollution Degradation"

  - Water contamination associated with human activities like urbanization, industrialization, and agricultural activities

## Literature Review

- The "Annual Review of Environment and Resources", section 'State of the World's Freshwater Ecosystems: Physical, Chemical, and Biological Changes'

  - The physical, chemical and biological changes in freshwater are associated with human factors

  - There is a lack of addressing the institutional aspects of water management.

## Research Goals

- Explore the correlation of aquatic life support and drinking water supply in the years 2006 and 2012

- Determine if there is a regional impact on the aquatic life support and drinking water supply correlation

- Determine how the length and area of the lakes changed over time

## Research Questions & Initial Hypotheses

- How are aquatic life support and drinking water supply correlated in the years 2006 and 2012?

  - H₀: There is no association between aquatic life support and drinking water supply among Connecticut lakes

  - Hₐ: There is an association between aquatic life support and drinking water supply among Connecticut lakes.

## Research Questions & Initial Hypotheses

- Do the regions where the lakes are located have an impact on the aquatic life support and drinking water supply correlation?

  - H₀: There is no association between the regions where the lakes are located and the impact on the aquatic life support and drinking water supply correlation

  - \- Hₐ: There is an association between the regions where the lakes are located and the impact on the aquatic life support and drinking water supply correlation.

## Research Questions & Initial Hypotheses

- How have the length and area changed over time?

  - H₀: There is no change in Lake size and length between 2006 and 2012

  - Hₐ: There is a significant difference in Lake size and length between 2006 and 2012

## Q1: Exploratory Data Analysis

```{r}
#| echo: false
#| message: false
#| warning: false

library(gt)

twelve_lakes <- read_csv("2012.csv")
six_lakes <- read_csv("2006.csv")
common_cols <- intersect(names(six_lakes), names(twelve_lakes))
six_lakes_aligned <- six_lakes |> select(all_of(common_cols))
twelve_lakes_aligned <- twelve_lakes |> select(all_of(common_cols))
total_lakes <- bind_rows(six_lakes_aligned, twelve_lakes_aligned) |>
  arrange(Cycle)

lakes_cleaned <- total_lakes |>
  filter(Cycle %in% c(2006, 2012)) |>
  mutate(
    aquatic_numeric = case_when(
      `Aquatic Life` == "Fully Supporting" ~ 5,
      `Aquatic Life` == "Not Supporting" ~ 4,
      `Aquatic Life` == "Not Assessed" ~ 3,
      `Aquatic Life` == "Not Evaluated In 2006" ~ 2,
      `Aquatic Life` == "Water Not Classified For Drinking Supply" ~ 1,
      TRUE ~ NA_real_
    ),
    drinking_numeric = case_when(
      `Drinking Water Supply` == "Fully Supporting" ~ 5,
      `Drinking Water Supply` == "Not Supporting" ~ 4,
      `Drinking Water Supply` == "Not Assessed" ~ 3,
      `Drinking Water Supply` == "Not Evaluated In 2006" ~ 2,
      `Drinking Water Supply` == "Water Not Classified For Drinking Supply" ~ 1,
      TRUE ~ NA_real_
    )
  ) |>
  filter(!is.na(aquatic_numeric) & !is.na(drinking_numeric))

lakes_wide <- lakes_cleaned |>
  pivot_wider(
    id_cols = `305B ID`,
    names_from = Cycle,
    values_from = c(aquatic_numeric, drinking_numeric)
  ) |>
  drop_na()

cor_matrix <- lakes_wide |>
  select(
    aquatic_numeric_2006,
    drinking_numeric_2006,
    aquatic_numeric_2012,
    drinking_numeric_2012
  ) |>
  cor(method = "spearman")

n_lakes <- nrow(lakes_wide)

cor_df <- as_tibble(cor_matrix, rownames = "Variable") |>
  rename(
    "Aquatic '06" = aquatic_numeric_2006,
    "Drinking '06" = drinking_numeric_2006,
    "Aquatic '12" = aquatic_numeric_2012,
    "Drinking '12" = drinking_numeric_2012
  )

cor_df |>
  gt() |>
  fmt_number(columns = -Variable, decimals = 2) |>
  tab_header(
    title = "Spearman Correlations: Aquatic Life & Drinking Water Support",
    subtitle = paste("Number of matched lakes (n):", n_lakes)
  ) |>
  tab_options(
    table.font.size = px(24),
    table.background.color = "transparent",
    heading.background.color = "transparent",
    table.font.color = "white"
  )

test_2006 <- cor.test(
  lakes_wide$aquatic_numeric_2006,
  lakes_wide$drinking_numeric_2006,
  method = "spearman"
)

test_2012 <- cor.test(
  lakes_wide$aquatic_numeric_2012,
  lakes_wide$drinking_numeric_2012,
  method = "spearman"
)

spearman_results <- tibble(
  Cycle = c("2006", "2012"),
  rho = c(test_2006$estimate, test_2012$estimate),
  S = c(test_2006$statistic, test_2012$statistic),
  p.value = c(test_2006$p.value, test_2012$p.value)
)

spearman_results |>
  gt() |>
  fmt_number(columns = c(rho, S), decimals = 3) |>
  fmt_number(columns = p.value, decimals = 4) |>
  tab_header(
    title = "Spearman Correlation Tests: Aquatic Life vs. Drinking Water"
  ) |>
  tab_options(
    table.background.color = "transparent",
    heading.background.color = "transparent",
    table.font.color = "white"
  )
```

## Q1: Exploratory Data Analysis

```{r}
lakes_cleaned |>
  ggplot(aes(x = drinking_numeric, y = aquatic_numeric, color = as.factor(Cycle))) +
  geom_jitter(alpha = 0.5, width = 0.18, height = 0.18) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 1.2) +
  scale_color_manual(values = c("2006" = "#56B4E9", "2012" = "#E69F00")) +
  labs(
    title = "Comparison of Trends: 2006 vs 2012",
    x = "Drinking Water Supply Score",
    y = "Aquatic Life Support Score",
    color = "Evaluation Year"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Q2: Exploratory Data Analysis

```{r}
total_lakes_regions <- lakes_cleaned |>
  separate_wider_regex(
    cols = Name,
    patterns = c(name = ".*?", " \\(", place = ".*?", "\\)"),
    too_few = "align_start"
  )

water_impact <- total_lakes_regions |>
  group_by(place, Cycle) |>
  summarise(
    mean_aquatic = mean(aquatic_numeric, na.rm = TRUE),
    mean_drinking = mean(drinking_numeric, na.rm = TRUE),
    .groups = "drop"
  )
district_1 <- water_impact |>
  filter(
    place %in% c(
      "Bristol", "Colebrook", "Hartland", "Manchester", "Newington", "Portland", "Winchester"
    ),
    Cycle %in% c(2006, 2012)
  )

district_2 <- water_impact |>
  filter(
    place %in% c("Ashford", "Bozrah", "Brooklyn", "Chester", "Colchester", "Columbia", "Coventry", "Durham", "East Haddam", "East Hampton", "East Lyme", "Griswold", "Haddam", "Hebron", "Killingly", "Lebanon", "Ledyard", "Mansfield", "Marlborough", "North Stonington", "Norwich", "Plainfield", "Preston", "Somers", "Sprague", "Sterling", "Thompson", "Tolland", "Union", "Vernon", "Voluntown", "Woodstock"),
    Cycle %in% c(2006, 2012)
  )

district_3 <- water_impact |>
  filter(
    place %in% c("Branford", "Derby", "Guilford", "Hamden", "New Haven", "North Branford", "Seymour", "Stratford", "Wallingford"),
    Cycle %in% c(2006, 2012)
  )

district_4 <- water_impact |>
  filter(
    place %in% c("Bridgeport", "Fairfield", "Greenwich", "Ridgefield", "Wilton"),
    Cycle %in% c(2006, 2012)
  )

district_5 <- water_impact |>
  filter(
    place %in% c("Canaan", "Cheshire", "Cornwall", "Danbury", "Goshen", "Kent", "Meriden", "New Fairfield", "Newtown", "Norfolk", "Plymouth", "Salisbury", "Sharon", "Simsbury", "Torrington", "Watertown", "Wolcott", "Woodbury"),
    Cycle %in% c(2006, 2012)
  )

cong_dist_master <- bind_rows(
  "district_1" = district_1,
  "district_2" = district_2,
  "district_3" = district_3,
  "district_4" = district_4,
  "district_5" = district_5,
  .id = "group"
)
c_d_plot <- cong_dist_master |>
  group_by(group, Cycle, place) |>
  summarise(
    avg_aquatic = mean(mean_aquatic, na.rm = TRUE),
    avg_drinking = mean(mean_drinking, na.rm = TRUE), .groups = "drop"
  )

cong_dist_plot1 <- ggplot(c_d_plot, aes(x = avg_drinking, y = avg_aquatic, color = group)) +
  geom_point(size = 3, alpha = 0.7, na.rm = TRUE) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2, na.rm = TRUE) +
  facet_wrap(~Cycle) +
  scale_color_discrete(
    labels = c(
      "district_1" = "District 1",
      "district_2" = "District 2",
      "district_3" = "District 3",
      "district_4" = "District 4",
      "district_5" = "District 5"
    )
  ) +
  labs(
    title = "Aquatic Life vs. Drinking Water Correlation by District",
    subtitle = "Varying line slopes indicate districts impact the correlation differently",
    x = "Mean Drinking Water Supply Score",
    y = "Mean Aquatic Life Support Score",
    color = "Congressional District"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    legend.position = "bottom"
  )
cong_dist_plot1
```

## Q2: Exploratory Data Analysis

```{r}
#| echo: false
#| message: false
#| warning: false

district_correlations <- c_d_plot |>
  group_by(group, Cycle) |>
  summarise(
    correlation = cor(
      avg_drinking,
      avg_aquatic,
      method = "spearman",
      use = "complete.obs"
    ),
    n = n(),
    .groups = "drop"
  )

district_correlations |>
  mutate(
    group = case_when(
      group == "district_1" ~ "District 1",
      group == "district_2" ~ "District 2",
      group == "district_3" ~ "District 3",
      group == "district_4" ~ "District 4",
      group == "district_5" ~ "District 5"
    )
  ) |>
  gt() |>
  cols_label(
    group = "District",
    Cycle = "Cycle",
    correlation = "Correlation",
    n = "N"
  ) |>
  fmt_number(columns = correlation, decimals = 3) |>
  sub_missing(columns = correlation, missing_text = "—") |>
  tab_header(
    title = "District-Level Spearman Correlations",
    subtitle = "Aquatic Life vs. Drinking Water Support, by Congressional District"
  ) |>
  tab_options(
    table.font.size = px(24),
    table.background.color = "transparent",
    heading.background.color = "transparent",
    table.font.color = "white",
    column_labels.font.weight = "bold"
  )
```

## Q3: Exploratory Data Analysis

```{r}
lakes_change <- six_lakes_aligned |>
  left_join(
    twelve_lakes_aligned,
    by = "305B ID",
    suffix = c("_2006", "_2012")
  )
lakes_change <- lakes_change |>
  select(Name_2006, Name_2012, Shape__Length_2006, Shape__Length_2012, Shape__Area_2006, Shape__Area_2012) |>
  mutate(length_ratio = Shape__Length_2012 / Shape__Length_2006, area_ratio = Shape__Area_2012 / Shape__Area_2006)

lakes_change <- lakes_change |>
  mutate(length_percent_change = (length_ratio - 1) * 100, area_percent_change = (area_ratio - 1) * 100)

ggplot(lakes_change, aes(x = length_percent_change)) +
  geom_histogram(bins = 30, fill = "steelblue") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Percent change in length", y = "Number of lakes", title = "Distribution of Changes in Lake Length from 2006 to 2012") +
  theme_minimal()
```

## Q3: Exploratory Data Analysis

```{r}
ggplot(lakes_change, aes(x = area_percent_change)) +
  geom_histogram(bins = 30, fill = "steelblue") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Percent change in area", y = "Number of lakes", title = "Distribution of Changes in Lake Area from 2006 to 2012") +
  theme_minimal()
```

## Q3: Exploratory Data Analysis

```{r}
#| echo: false
#| message: false
#| warning: false

length_test <- t.test(lakes_change$Shape__Length_2012, lakes_change$Shape__Length_2006, paired = TRUE)
area_test <- t.test(lakes_change$Shape__Area_2012, lakes_change$Shape__Area_2006, paired = TRUE)

t_test_results <- tibble(
  measure = c("Shape Length", "Shape Area"),
  estimate = c(length_test$estimate, area_test$estimate),
  statistic = c(length_test$statistic, area_test$statistic),
  p.value = c(length_test$p.value, area_test$p.value),
  conf.low = c(length_test$conf.int[1], area_test$conf.int[1]),
  conf.high = c(length_test$conf.int[2], area_test$conf.int[2])
)

t_test_results |>
  gt() |>
  fmt_number(columns = -measure, decimals = 3) |>
  tab_header(
    title = "Paired t-tests: 2006 vs. 2012"
  ) |>
  tab_options(
    table.background.color = "transparent",
    heading.background.color = "transparent",
    table.font.color = "white"
  )
```

## Methods, Modeling, Results

- Question: How are aquatic life support and drinking water supply correlated in the years 2006 and 2012?

<!-- -->

- (H₀): There is no association between aquatic life support and drinking water supply (rejected)

- Visualization: Scatterplot showed a weak positive relationship

- Spearman correlation, 2006: ρ = 0.111; Spearman correlation, 2012: ρ = 0.116

- Assumption: Variables were ordinal rather than continuous

## Methods, Modeling, Results

- Question: Do the congressional districts where the lakes are located have an impact on the aquatic life support and drinking water supply correlation?

- H₀: Congressional district has no effect on the correlation between aquatic life support and drinking water supply

- Visualization: Scatterplots for each congressional district for 2006 and 2012

- Spearman correlations for districts 2–5 showed weak positive relationships, Correlations could not be calculated for District 1

- Assumption: Ordinal data and monotonic relationship

## Methods, Modeling, Results

- Question: How have the length and area changed over time?

- H₀: There was no change in lake length and area between 2006 and 2012

- Visualization: Histograms of percent change for length and area

- Paired t-test lake area:*t* = 0.134, *p* = 0.893; Paired t-test lake length: *t* = −2.085, *p* = 0.039

- Assumption: Paired observations and normality of distribution

## Data Driven Questions

- Observation: A significant decrease in lake length was observed from 2006 to 2012

- Question: How does drinking water supply affect lake length from 2006 to 2012?

- Question: Does the reduction in lake length from 2006 to 2012 affect aquatic life support? 

## Data Driven Questions

- Observation: The data did not suggest impact of congressional district on aquatic life support and drinking water supply

- Question: How is recreation affected by congressional district?

## Data Driven Questions

- Observation: Aquatic life support and drinking water supply had a very weak positive correlation

- Question: What characteristics are associated with aquatic life as opposed to drinking water?

- Question: What is the relationship between recreation and aquatic life support?

## Discussion: Question 1

There is a slight visual positive correlation between Aquatic Life Support score and Drinking Water Supply score between 2006 and 2012 with 2012 having a greater positive correlation. As Drinking Water Supply score increased, Aquatic Life Support score increased at a greater rate in 2012 than it did in 2006. Based on the Spearman’s Correlation Test, the p-value in 2006 of 0.1776 and the p-value in 2012 of 0.1568, there is no evidence of an association between aquatic life support and drinking water supply in CT lakes.

## Discussion: Question 2

There is a slight visual positive correlation between Aquatic Life Support score and Drinking Water Supply score between both years and in all districts. While there were slight changes between the years of 2006 and 2012, the major differences were between districts. Notably, District 4 had the lowest Mean Aquatic Life Support score out of all the other districts in both years that were observed and District 2 had the lowest Mean Drinking Water score out of all the districts in both years. There is a very slight association between Aquatic Life Support score and region.

## Discussion: Question 3

For question 3 regarding the change in length of the lakes from 2006 to 2012, there was very little change. The peak of the histogram, about 37 lakes, is just below 0.000, with the far left being -0.0125% change in length, and the far right being about +0.006% change in length between the years. The change in area of lakes was also very minimal. The peak of the histogram showing percent area change is almost 80 lakes at 0.00% change in area between 2006 and 2012. The far left margin is about -0.0225% change in area, and the far right margin is about +0.0175% change in area. The results of the paired t-test show that there was not a significant change in lake length and area between 2006 and 2012. While the p-value for length was significant at 0.039, the p-value for area was large at 0.893, making the null hypothesis true.

## References

- Bashir, I., Lone, F. A., Bhat, R. A., Mir, S. A., Dar, Z. A., & Dar, S. A. (2020). Concerns and Threats of Contamination on Aquatic Ecosystems. Bioremediation and Biotechnology: Sustainable Approaches to Pollution Degradation, 1–26. <https://doi.org/10.1007/978-3-030-35691-0_1>

- Stephen R. Carpenter, Emily H. Stanley, M. Jake Vander Zanden. 2011. State of the World's Freshwater Ecosystems: Physical, Chemical, and Biological Changes. Annual Review of Environment and Resources 36:75-99. <https://doi.org/10.1146/annurev-environ-021810-094524>

##  {.center}

::: r-fit-text
Thank You!
:::
